Nodal Discontinuous Galerkin Method for the Two-Moment Model of Neutrino Transport

### 2017-07-26 Documents/M1/dgM1Curvilinear.tex
Added discussion of full interaction solver.  Added HLL flux expression.

To connect the elements in Eq.~\eqref{eq:semidiscreteDG}, $\widehat{\bcF}^{i}(\bcM_{\DG})$ is a numerical flux approximating the flux on the $i$th surface of $\bK$.  
For this purpose we define the numerical flux function $\vect{f}^{i}$, which evaluates the numerical flux given values from both sides of an element interface; i.e.,
\begin{equation}
  \widehat{\bcF}^{i}(\bcM_{\DG})=\vect{f}^{i}(\bcM_{\DG}(x^{i,-},\tilde{\bx}^{i}),\bcM_{\DG}(x^{i,+},\tilde{\bx}^{i})),
\end{equation}
where superscripts $-/+$, e.g., in the arguments of $\bcM_{\DG}(x^{i,-/+},\tilde{\bx}^{i})$, indicate that the function is evaluated to the immediate left/right of $x^{i}$.  
For example, the simple Lax-Friedrichs (LF) flux, which we have implemented, is given by
\begin{equation}
  \vect{f}^{\mbox{\tiny LF},i}(\bcM^{-},\bcM^{+})
  =\f{1}{2}\,\big(\,\bcF^{i}(\bcM^{-})+\bcF^{i}(\bcM^{+})-\alpha^{i}\,(\,\bcM^{+}-\bcM^{-}\,)\,\big),
\end{equation}
where $\alpha^{i}=||\mbox{eig}\big(\partial\bcF^{i}/\partial\bcM\big)||_{\infty}$ is the largest eigenvalue of the flux jacobian.  
For classical neutrinos, which propagate at the speed of light, we take $\alpha^{i}=1$ (i.e., the global LF flux).  
We have also implemented the Harten-Lax-van Leer (HLL) flux
\begin{equation}
  \vect{f}^{\mbox{\tiny HLL},i}(\bcM^{-},\bcM^{+})
  =
\end{equation}

@@

In the following subsections we separately give the collision terms for the two-moment model by taking angular moments of Eq.~\eqref{eq:homogeneousBoltzmann}.  
(The total collision term in the two-moment model is then given by the sum of the separate terms.)  
The opacities are computed using the SFHo equation of state \citep{steiner_etal_2013}, and tabulated for densities $\rho\in[1.66\times10^{4},3.16\times10^{15}]$~g~cm$^{-3}$, temperatures $T\in[1.16\times10^{9},1.84\times10^{12}]$~K, and electron fractions $Y_{e}\in[0.01,0.6]$.  
The neutrino energies are in the range $\epsilonNu\in[0.1,300]$~MeV.  
We use logarithmic grids to cover the range in density $N_{\rho}=308$, temperature ($N_{T}=129$), and the neutrino energy ($N_{\epsilonNu}=40$), and a linear grid to cover the range in electron fraction ($N_{Y_{e}}=60$).  
The neutrino electron scattering rates are tabulated in terms of neutrino energies ($\epsilonNu$ and $\epsilonNu'$), temperature, and $\eta=\mu_{\nu}/(k_{\mbox{\tiny B}}T)$.  
Here we use a logarithmic grid in $\eta\in[1.0\times10^{-8},2.5\times10^{3}]$~MeV, using $N_{\eta}=100$ points.

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{cc}
      \includegraphics[scale=0.4]{./Figures/MilneProblem1D_n20.png} &
      \includegraphics[scale=0.4]{./Figures/MilneProblem1D_n20_EddingtonFactors.png} \\
      \includegraphics[scale=0.4]{./Figures/MilneProblem1D_n15.png} &
      \includegraphics[scale=0.4]{./Figures/MilneProblem1D_n30.png}
    \end{tabular}
  \end{center}
  \caption{Results of the Milne problem obtained with the two-moment model (solid lines), obtained with the DG(2)+SIRK2 scheme, compared with solutions provided by \citep{hummerRybicki_1971} (boxes).  See text for details.}
  \label{fig:milneProblem1D}
\end{figure}

@@

Electron capture on nucleons and nuclei results in changes to the electron fraction and internal energy of the fluid given by
\begin{align}
  d_{t}Y_{e}
  &=-\f{m_{\mbox{\tiny B}}}{\rho}\int_{\bbR^{+}}\tilde{\chi}\,\big(\cJ_{0}-\cJ\big)\,dV_{\epsilonNu}, \label{eq:electronFractionEmAb} \\
  d_{t}\epsilon
  &=-\f{1}{\rho}\int_{\bbR^{+}}\tilde{\chi}\,\big(\cJ_{0}-\cJ\big)\,\epsilonNu\,dV_{\epsilonNu}, \label{eq:internalEnergyEmAb}
\end{align}
where $m_{\mbox{\tiny B}}$ is the baryon mass and $\epsilon$ is the specific internal energy.  
In each implicit step in the SIRK2 scheme discussed in Section~\ref{sec:semiImplicitTime} we solve Eqs.~\eqref{eq:electronFractionEmAb}-\eqref{eq:internalEnergyEmAb} coupled to the collision part of the moment equations.  
\begin{align}
  Y_{e}^{n+1}-Y_{e}^{n}
  &=-\f{m_{\mbox{\tiny B}}}{\rho}\sum_{j=1}^{M^{\epsilonNu}}W_{j}^{(2)}\,\gamma_{j}^{n}\,\big(\cJ_{0,j}^{n+1}-\cJ_{j}^{n+1}\big), \label{eq:electronFractionEmAbDiscrete0} \\
  \epsilon^{n+1}-\epsilon^{n}
  &=-\f{1}{\rho}\sum_{j=1}^{M^{\epsilonNu}}W_{j}^{(3)}\,\gamma_{j}^{n}\,\big(\cJ_{0,j}^{n+1}-\cJ_{j}^{n+1}\big), \label{eq:internalEnergyEmAbDiscrete0}
\end{align}
where $W_{j}^{(2,3)}=w_{q}\,\epsilonNu_{i,q}^{2,3}\,\Delta\epsilonNu_{i}$ and $\gamma_{j}^{n}=\dt\,\tilde{\chi}_{j}^{n}$.  
Note that the absorption opacity is defined at the known time level $t^{n}$, which avoids computation of derivatives with respect to temperature and electron fraction.  
The radiation energy density is updated from
\begin{equation}
  \cJ_{j}^{n+1}=\f{\cJ_{j}^{n}+\gamma_{j}^{n}\,\cJ_{0,j}^{n+1}}{1+\gamma_{j}^{n}}, 
  \label{eq:momentEquationEnergyEmAbDiscrete}
\end{equation}
which, when inserted into Eqs.~\eqref{eq:electronFractionEmAbDiscrete0}-\eqref{eq:internalEnergyEmAbDiscrete0}, results in a nonlinear system for the unknowns $Y^{n+1}$ and $\epsilon^{n+1}$
\begin{align}
  Y_{e}^{n+1}-Y_{e}^{n}
  &=-\f{m_{\mbox{\tiny B}}}{\rho}\sum_{j=1}^{M^{\epsilonNu}}W_{j}^{(2)}\,\tilde{\gamma}_{j}^{n}\,\big(\cJ_{0,j}^{n+1}-\cJ_{j}^{n}\big), \label{eq:electronFractionEmAbDiscrete} \\
  \epsilon^{n+1}-\epsilon^{n}
  &=-\f{1}{\rho}\sum_{j=1}^{M^{\epsilonNu}}W_{j}^{(3)}\,\tilde{\gamma}_{j}^{n}\,\big(\cJ_{0,j}^{n+1}-\cJ_{j}^{n}\big), \label{eq:internalEnergyEmAbDiscrete}
\end{align}
where $\tilde{\gamma}_{j}^{n}=\gamma_{j}^{n}/(1+\gamma_{j}^{n})$.  
Since the equilibrium density $\cJ_{0,j}$ is a nonlinear function of $Y_{e}$ and $\epsilon$, Eqs.~\eqref{eq:electronFractionEmAbDiscrete}-\eqref{eq:internalEnergyEmAbDiscrete} are solved using Newton's method.  
Once $Y_{e}^{n+1}$ and $\epsilon^{n+1}$ are obtained, the radiation density is obtained from Eq.~\eqref{eq:momentEquationEnergyEmAbDiscrete}, while the flux is given by
\begin{equation}
  \bcH_{j}^{n+1}=\bcH_{j}^{n}/(1+\gamma_{j}^{n}).  
  \label{eq:momentEquationMomentumEmAbDiscrete}
\end{equation}

@@

>>>>

To connect the elements in Eq.~\eqref{eq:semidiscreteDG}, $\widehat{\bcF}^{i}(\bcM_{\DG})$ is a numerical flux approximating the flux on the $i$th surface of $\bK$.  
For this purpose we define the numerical flux function $\vect{f}^{i}$, which evaluates the numerical flux given values from both sides of an element interface; i.e.,
\begin{equation}
  \widehat{\bcF}^{i}(\bcM_{\DG})=\vect{f}^{i}(\bcM_{\DG}(x^{i,-},\tilde{\bx}^{i}),\bcM_{\DG}(x^{i,+},\tilde{\bx}^{i})),
\end{equation}
where superscripts $-/+$, e.g., in the arguments of $\bcM_{\DG}(x^{i,-/+},\tilde{\bx}^{i})$, indicate that the function is evaluated to the immediate left/right of $x^{i}$.  
For example, the simple Lax-Friedrichs (LF) flux, which we have implemented, is given by
\begin{equation}
  \vect{f}^{\mbox{\tiny LF},i}(\bcM^{-},\bcM^{+})
  =\f{1}{2}\,\big(\,\bcF^{i}(\bcM^{-})+\bcF^{i}(\bcM^{+})-\alpha^{i}\,(\,\bcM^{+}-\bcM^{-}\,)\,\big),
\end{equation}
where $\alpha^{i}=||\mbox{eig}\big(\partial\bcF^{i}/\partial\bcM\big)||_{\infty}$ is the largest eigenvalue of the flux jacobian.  
For classical neutrinos, which propagate at the speed of light, we can take $\alpha^{i}=1$ (i.e., the global LF flux).  
We have also implemented the Harten-Lax-van Leer (HLL) flux \citep{harten_etal_1983}
\begin{equation}
  \vect{f}^{\mbox{\tiny HLL},i}(\bcM^{-},\bcM^{+})
  =\f{\alpha^{+,i}\,\bcF^{i}(\bcM^{-})+\alpha^{-,i}\,\bcF^{i}(\bcM^{+})-\alpha^{-,i}\alpha^{+,i}(\,\bcM^{+}-\bcM^{-}\,)}{\alpha^{-,i}+\alpha^{+,i}},
\end{equation}
where $\alpha^{-,i}$ and $\alpha^{+,i}$ are estimates for the largest (in absolute value) wave speed for left-moving and right-moving waves, respectively.

@@

In the following subsections we separately give the collision terms for the two-moment model by taking angular moments of Eq.~\eqref{eq:homogeneousBoltzmann}.  
(The total collision term in the two-moment model is then given by the sum of the separate terms.)  
The opacities are computed using the SFHo equation of state \citep{steiner_etal_2013}, and tabulated for densities $\rho\in[1.66\times10^{4},3.16\times10^{15}]$~g~cm$^{-3}$, temperatures $T\in[1.16\times10^{9},1.84\times10^{12}]$~K, and electron fractions $Y_{e}\in[0.01,0.6]$.  
The neutrino energies are in the range $\epsilonNu\in[0.1,300]$~MeV.  
We use logarithmic grids to cover the range in density $N_{\rho}=308$, temperature ($N_{T}=161$), and the neutrino energy ($N_{\epsilonNu}=40$), and a linear grid to cover the range in electron fraction ($N_{Y_{e}}=60$).  
The neutrino electron scattering rates are tabulated in terms of neutrino energies ($\epsilonNu$ and $\epsilonNu'$), temperature, and $\eta=\mu_{\nu}/(k_{\mbox{\tiny B}}T)$.  
Here we use a logarithmic grid in $\eta\in[1.0\times10^{-8},2.5\times10^{3}]$~MeV, using $N_{\eta}=100$ points.

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{cc}
      \includegraphics[scale=0.4]{./Figures/MilneProblem1D_n20.png} &
      \includegraphics[scale=0.4]{./Figures/MilneProblem1D_n20_EddingtonFactors.png} \\
      \includegraphics[scale=0.4]{./Figures/MilneProblem1D_n15.png} &
      \includegraphics[scale=0.4]{./Figures/MilneProblem1D_n30.png}
    \end{tabular}
  \end{center}
  \caption{Results of the Milne problem obtained with the two-moment model (solid lines), using the DG(2)+SIRK2 scheme, compared with solutions provided by \citep{hummerRybicki_1971} (boxes).  See text for details.}
  \label{fig:milneProblem1D}
\end{figure}

@@

Electron capture on nucleons and nuclei results in changes to the electron fraction and internal energy of the fluid given by
\begin{align}
  d_{t}Y_{e}
  &=-\f{m_{\mbox{\tiny B}}}{\rho}\int_{\bbR^{+}}\tilde{\chi}\,\big(\cJ_{0}-\cJ\big)\,dV_{\epsilonNu}, \label{eq:electronFractionEmAb} \\
  d_{t}\epsilon
  &=-\f{1}{\rho}\int_{\bbR^{+}}\tilde{\chi}\,\big(\cJ_{0}-\cJ\big)\,\epsilonNu\,dV_{\epsilonNu}, \label{eq:internalEnergyEmAb}
\end{align}
where $m_{\mbox{\tiny B}}$ is the baryon mass and $\epsilon$ is the specific internal energy.  
In each implicit step in the SIRK2 scheme discussed in Section~\ref{sec:semiImplicitTime} we solve Eqs.~\eqref{eq:electronFractionEmAb}-\eqref{eq:internalEnergyEmAb} coupled to the collision part of the moment equations.  
\begin{align}
  Y_{e}^{n+1}-Y_{e}^{n}
  &=-\f{m_{\mbox{\tiny B}}}{\rho}\sum_{j=1}^{M^{\epsilonNu}}W_{j}^{(2)}\,\gamma_{j}^{n}\,\big(\cJ_{0,j}^{n+1}-\cJ_{j}^{n+1}\big), \label{eq:electronFractionEmAbDiscrete0} \\
  \epsilon^{n+1}-\epsilon^{n}
  &=-\f{1}{\rho}\sum_{j=1}^{M^{\epsilonNu}}W_{j}^{(3)}\,\gamma_{j}^{n}\,\big(\cJ_{0,j}^{n+1}-\cJ_{j}^{n+1}\big), \label{eq:internalEnergyEmAbDiscrete0}
\end{align}
where $W_{j}^{(2,3)}=w_{q}\,\epsilonNu_{i,q}^{2,3}\,\Delta\epsilonNu_{i}$ and $\gamma_{j}^{n}=\dt\,\tilde{\chi}_{j}^{n}$.  
Note that the absorption opacity is defined at the known time level $t^{n}$, which avoids computation of derivatives with respect to temperature and electron fraction.  
The radiation energy density is updated from
\begin{equation}
  \cJ_{j}^{n+1}=\f{\cJ_{j}^{n}+\gamma_{j}^{n}\,\cJ_{0,j}^{n+1}}{1+\gamma_{j}^{n}}, 
  \label{eq:momentEquationEnergyEmAbDiscrete}
\end{equation}
which, when inserted into Eqs.~\eqref{eq:electronFractionEmAbDiscrete0}-\eqref{eq:internalEnergyEmAbDiscrete0}, results in a nonlinear system for the unknowns $Y^{n+1}$ and $\epsilon^{n+1}$
\begin{align}
  Y_{e}^{n+1}-Y_{e}^{n}
  &=-\f{m_{\mbox{\tiny B}}}{\rho}\sum_{j=1}^{M^{\epsilonNu}}W_{j}^{(2)}\,\tilde{\gamma}_{j}^{n}\,\big(\cJ_{0,j}^{n+1}-\cJ_{j}^{n}\big), \label{eq:electronFractionEmAbDiscrete} \\
  \epsilon^{n+1}-\epsilon^{n}
  &=-\f{1}{\rho}\sum_{j=1}^{M^{\epsilonNu}}W_{j}^{(3)}\,\tilde{\gamma}_{j}^{n}\,\big(\cJ_{0,j}^{n+1}-\cJ_{j}^{n}\big), \label{eq:internalEnergyEmAbDiscrete}
\end{align}
where $\tilde{\gamma}_{j}^{n}=\gamma_{j}^{n}/(1+\gamma_{j}^{n})$.  
Since the equilibrium density $\cJ_{0,j}$ is a nonlinear function of $Y_{e}$ and $\epsilon$, Eqs.~\eqref{eq:electronFractionEmAbDiscrete}-\eqref{eq:internalEnergyEmAbDiscrete} are solved using Newton's method.  
Once $Y_{e}^{n+1}$ and $\epsilon^{n+1}$ are obtained, the radiation density is obtained from Eq.~\eqref{eq:momentEquationEnergyEmAbDiscrete}, while the flux is given by
\begin{equation}
  \bcH_{j}^{n+1}=\bcH_{j}^{n}/(1+\gamma_{j}^{n}).  
  \label{eq:momentEquationMomentumEmAbDiscrete}
\end{equation}
When elastic scattering is included, the above description remains unchanged, except for $\gamma_{j}^{n}\to\dt\big(\tilde{\chi}_{j}^{n}+\sigma_{j}^{n}\big)$ in Eq.~\eqref{eq:momentEquationMomentumEmAbDiscrete}.

@@

With the complete interaction set considered in this paper (i.e., emission and absorption, elastic scattering, and NES), the electron fraction and fluid internal energy evolve according to
\begin{align}
  d_{t}Y_{e}
  &=-\f{m_{\mbox{\tiny B}}}{\rho}\int_{\bbR^{+}}\tilde{\chi}\,\big(\cJ_{0}-\cJ\big)\,dV_{\epsilonNu}, \\
  \rho\,d_{t}\epsilon
  &=-\int_{\bbR^{+}}\tilde{\chi}\,\big(\cJ_{0}-\cJ\big)\,\epsilonNu\,dV_{\epsilonNu} \nonumber \\
  &\hspace{12pt}
  -\int_{\bbR^{+}}\int_{\bbR^{+}}\big(4\pi-\cJ(\epsilonNu)\big)R^{\In}(\epsilonNu,\epsilonNu')\,\cJ(\epsilonNu')\,dV_{\epsilonNu'}\,\epsilonNu\,dV_{\epsilonNu} \nonumber \\
  &\hspace{12pt}
  +\int_{\bbR^{+}}\int_{\bbR^{+}}\cJ(\epsilonNu)\,R^{\Out}(\epsilonNu,\epsilonNu')\,\big(4\pi-\cJ(\epsilonNu)\big)\,dV_{\epsilonNu'}\,\epsilonNu\,dV_{\epsilonNu},
\end{align}
and the conserved quantities are the total lepton number
\begin{equation}
  d_{t}\big(\,\f{\rho}{m_{\mbox{\tiny B}}}Y_{e}+\int_{\bbR^{+}}\cJ\,dV_{\epsilonNu}\,\big)=0,
  \label{eq:leptonConservationFull}
\end{equation}
and the total (fluid plus radiation) energy
\begin{equation}
  d_{t}\big(\,\rho\,\epsilon+\int_{\bbR^{+}}\cJ\,\epsilonNu\,dV_{\epsilonNu}\,\big)=0.  
  \label{eq:energyConservationFull}
\end{equation}

In the solution to the coupled system, we use Eqs.~\eqref{eq:leptonConservationFull} and \eqref{eq:energyConservationFull} together with the equation for the neutrino number density to form a system of nonlinear equations for the solution vector $\bU=(\bcJ,Y_{e},\epsilon)^{\mbox{\tiny T}}$, i.e.,
\begin{equation}
  \bF(\bU^{n+1})=0,
  \label{eq:nonlinearSystemFull}
\end{equation}
where, for $j=1,\ldots,M^{\epsilonNu}$,
\begin{align}
  F_{j}(\bU^{n+1})
  &=\big(\cJ_{j}^{n+1}-\cJ_{j}^{n}\big)-\dt\,\tilde{\chi}_{j}^{n}\big(\,\cJ_{0,j}^{n+1}-\cJ_{j}^{n+1}\,\big) \nonumber \\
  &\hspace{12pt}
  -\dt\,\Big\{\,
    \big(4\pi-\cJ_{j}^{n+1}\big)\sum_{k=1}^{M^{\epsilonNu}}W_{k}^{(2)}\,R_{jk}^{\In,n}\,\cJ_{k}^{n+1}
    -\cJ_{j}^{n+1}\sum_{k=1}^{M^{\epsilonNu}}W_{k}^{(2)}\,R_{jk}^{\Out,n}\,\big(4\pi-\cJ_{k}^{n+1}\big)\,\Big\}.
\end{align}
The electron fraction is constrained by
\begin{equation}
  F_{M^{\epsilonNu}+1}(\bU^{n+1})
  =Y_{e}^{n+1}-Y_{e}^{n}+\f{m_{\mbox{\tiny B}}}{\rho}\sum_{k=1}^{M^{\epsilonNu}}W_{k}^{(2)}\big(\,\cJ_{k}^{n+1}-\cJ_{k}^{n}\,\big),
\end{equation}
while the internal energy is constrained by
\begin{equation}
  F_{M^{\epsilonNu}+2}(\bU^{n+1})
  =\epsilon^{n+1}-\epsilon^{n}+\f{1}{\rho}\sum_{k=1}^{M^{\epsilonNu}}W_{k}^{(3)}\big(\,\cJ_{k}^{n+1}-\cJ_{k}^{n}\,\big).  
\end{equation}
The system given by Eq.~\eqref{eq:nonlinearSystemFull} is solved using Newton's method, where $\bU^{n}$ is taken as an initial guess.  
The dense Jacobian matrix is computed analytically, except for derivatives of the equilibrium density $\cJ_{0}$, which are computed numerically by taking derivatives of table interpolation functions.  
The radiation flux is subsequently obtained from
\begin{equation}
  \bcH_{j}^{n+1}-\bcH_{j}^{n}
  =-\dt\,\big(\,\tilde{\chi}_{j}^{n}+\sigma_{j}^{n}+\sigma_{\NES,j}^{n}\,\big)\,\bcH_{j}^{n+1},
\end{equation}
where
\begin{equation}
  \sigma_{\NES,j}^{n}
  =\sum_{k=1}^{M^{\epsilonNu}}W_{k}^{(2)}\big[\,R_{jk}^{\In,n}\,\cJ_{k}^{n+1}+R_{jk}^{\Out,n}\big(4\pi-\cJ_{k}^{n+1}\big)\,\big].
\end{equation}
